RegressionPvNP

### 2017-05-10 RegressionPvNP.Rmd
…

---
title: "RegressionPvNP"
author: "Sarah Killcoyne"
date: "31 March 2017"
output: 
html_document: 
fig_caption: yes
fig_height: 10
fig_width: 10
toc: yes
---

@@

## cv.glmnet Ridge vs Lasso

```{r cvglmnetPNP, echo=F, warning=F, message=F, eval=F}
#With all `r ncol(dysplasia.df)` regions of the genome, first check if/which regression may be appropriate. It appears that pure lasso may provide the best balance for number of features with non-zero coefficients.

@@

      fit.e[[length(fit.e)+1]] = fit
    }
    cv.pred = rbind(cv.pred, cv.class)    
    cv.binomial.deviance = rbind(cv.binomial.deviance, deviance)
  }

  # Not really sure what to do with the binomial deviance now...
  df = cbind.data.frame('lambda.at'=1:ncol(cv.pred),
                        'mean'= colMeans(cv.pred), 
                        'sme'= apply(cv.pred, 2, sd)/sqrt(nrow(cv.pred)), 
                        'sd'= apply(cv.pred, 2, sd), 
                        'lambda'= lambda,
                        'log.lambda' = log(lambda),
                        'mean.b.dev' = colMeans(cv.binomial.deviance),
                        'sd.b.dev' = apply(cv.binomial.deviance, 2,sd),
                        'sme.b.dev' = apply(cv.binomial.deviance, 2, sd)/sqrt(nrow(cv.binomial.deviance)),
                        'lambda.min' = F, 'lambda.1se' = F)
  #df$lambda.1se = FALSE
  #df$`lambda-1se` = FALSE

@@

```{r xvalpt, message=F, warning=F, echo=T, fig.height=12, fig.width=12}
performance.plots = list(); coefs = list(); deviance.plots = list()
plots = list()
folds = 10; splits=5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
  autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

@@

  plots[[as.character(a)]] = arrangeGrob(cv.patient$plot+ggtitle('CV Performance'), cv.patient$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
}

@@

@@

Values that are closer to full lasso (alpha=1) are pretty similar, however at full lasso we see fewer features than if we regularize it at 0.8.

`r pander(as.data.frame(coef.stable[['1']])/(folds*splits), justify='left', caption='Percentage of features from the glm fitted with all data. Numbers indicate the total number of folds that feature was identified in.')`

They also share all of the non-zero coefficients (0.8 identifies the most non-zero coefs `r max(sapply(coef.stable, length))`).

`r pander(cfs, justify='left')`

#### Did I understand correctly?

In order to predict, my trained model is then my original model, but using the optimal lambda value?
This isn't quite correct because the model was trained including these patients, just an example.

```{r}
fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=200, family='binomial') # all patients
cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=fit0$lambda, pts=pts, a=1, nfolds=20, splits=5, fit=fit0)

# Progressor, minus the HGD sample(s)
cbind.data.frame(
  predict(fit0, dysplasia.df[patient.data$PR1_HIN_042$info$Samplename[-(9:10)],], s=cv.patient$lambda.1se, type='response'),
  predict(fit0, dysplasia.df[patient.data$PR1_HIN_042$info$Samplename[-(9:10)],], s=cv.patient$lambda.1se, type='class'))

# Non-P, all correctly predicted
cbind.data.frame(predict(fit0, dysplasia.df[patient.data$AD0610$info$Samplename,], s=cv.patient$lambda.1se, type='response'),
                 predict(fit0, dysplasia.df[patient.data$AD0610$info$Samplename,], s=cv.patient$lambda.1se, type='class'))

```{r noHGD, echo=T, message=F, warning=F, fig.height=12, fig.width=12}
`%nin%` <- Negate(`%in%`)

# No HGD/IMC
samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC'))$Samplename)

plots = list()
coefs = list()
for (a in alpha.values) {
  fit1 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  autoplot(fit1, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
  lambda1 <- fit1$lambda

  cv.nohgd = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=lambda1, pts=subset(pts, Samplename %in% samples), a=a, nfolds=folds, splits=splits, fit=fit1)

  plots[[as.character(a)]] = arrangeGrob(cv.patient$plot+ggtitle('CV Performance'), cv.patient$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

  coef.1se = as.data.frame(non.zero.coef(fit1, cv.nohgd$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nohgd$non.zero.cf)
}

do.call(grid.arrange, c(plots, top="No HGD/IMC samples", ncol=2))

@@

`r pander(as.data.frame(coef.stable[['1']])/(folds*splits), justify='left', caption='Percentage of features from the glm fitted with all data. Numbers indicate the total number of folds that feature was identified in.')`

@@

`r pander(cfs, justify='left')`

```{r noLGD, echo=T, message=F, warning=F, fig.height=8, fig.width=8}
# No LGD
samples = intersect(rownames(dysplasia.df), c(subset(info, Status == 'NP')$Samplename, subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD') & Status == 'P')$Samplename))

plots = list(); coefs = list()
for (a in alpha.values) {
  fit2 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  autoplot(fit2, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
  lambda2 <- fit2$lambda

  cv.nolgd = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=lambda2, pts=subset(pts, Samplename %in% samples), a=a, nfolds=folds, splits=splits, fit=fit2)

  plots[[as.character(a)]] = arrangeGrob(cv.nolgd$plot+ggtitle('CV Performance'), cv.nolgd$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

  coef.1se = as.data.frame(non.zero.coef(fit2, cv.nolgd$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nolgd$non.zero.cf)
}

for (i in names(coef.stable)) 
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1
```

`r pander(as.data.frame(coef.stable[['1']])/(folds*splits), justify='left', caption='Percentage of features from the glm fitted with all data. Numbers indicate the total number of folds that feature was identified in.')`

`r pander(cfs, justify='left')`

@@

```{r, echo=T, message=F, warning=F, eval=F}
m = median(subset(sum.patient.data, Status == 'P')$total.samples)

@@

few.plots = list()
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, nlambda=500, family='binomial') # all patients
  cv.back.few = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=a, fit=fit0, minR=0.1)
  #few.plots[[as.character(a)]] = cv.back.few$plot + ggtitle(paste('alpha=',a,sep=''))

@@

  cv.back.many = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=a, fit=fit0)
  #many.plots[[as.character(a)]] = cv.back.many$plot + ggtitle(paste('alpha=',a,sep=''))

@@

  # Fit generated on all samples, including HGD
  a = 1
  fitLOO <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  cv = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fitLOO$lambda, 
                                pts=subset(pts, Samplename %in% samples), a=a, nfolds=10, splits=5, fit=fitLOO)
  plots[[pt]] = cv$plot + ggtitle(pt)
  fits[[pt]] = cv

@@

    coef.1se = as.data.frame(non.zero.coef(fit1, cv$lambda.1se))
    coefs[[pt]] = coef.stability(coef.1se, cv$non.zero.cf)

    pg.samp[[pt]]$Prediction = predict(fitLOO, dysplasia.df[pg.samp[[pt]]$Samplename,], s=cv$lambda.1se, type='response')[,1]
  } else {
    warning(paste("Patient", pt, "did not have a 1se"))
  }
}
save(plots, performance.at.1se, coefs, fits, file='/tmp/loo.Rdata')
}

@@

When predicting each sample individually `r length(pg[pg == 1])` patients were correctly predicted in all samples. The remaining patients were predicted in 80% or fewer samples.

@@

```{r timepoints2, echo=T, message=F, warning=F, eval=F, fig.height=400, fig.width=100}
# For all pre-HGD/IMC/LGD in progressors
pre.hgd = sapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  e = subset(pt, Pathology %nin% c('HGD', 'IMC', 'LGD'))
  e$Prediction
  sum(as.integer(e$Prediction > 0.5))/nrow(e)
})

@@

>>>>

---
title: "RegressionPvNP"
author: "Sarah Killcoyne"
date: "31 March 2017"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 10
    fig_width: 10
    number_sections: yes
    toc: yes
---

@@

```{r cvglmnetPNP, echo=F, warning=F, message=F, eval=F}
## cv.glmnet Ridge vs Lasso
#With all `r ncol(dysplasia.df)` regions of the genome, first check if/which regression may be appropriate. It appears that pure lasso may provide the best balance for number of features with non-zero coefficients.

@@

      fit.e[[length(fit.e)+1]] = fitCV
    }
    cv.pred = rbind(cv.pred, cv.class)    
    cv.binomial.deviance = rbind(cv.binomial.deviance, deviance)
  }

  # Not really sure what to do with the binomial deviance now...
  df = cbind.data.frame('lambda.at'=1:ncol(cv.pred),
                        'mean'= colMeans(cv.pred), 
                        'sme'= apply(cv.pred, 2, sd)/sqrt(nrow(cv.pred)), 
                        'sd'= apply(cv.pred, 2, sd), 
                        'lambda'= lambda,
                        'log.lambda' = log(lambda),
                        'mean.b.dev' = colMeans(cv.binomial.deviance),
                        'sd.b.dev' = apply(cv.binomial.deviance, 2,sd),
                        'sme.b.dev' = apply(cv.binomial.deviance, 2, sd)/sqrt(nrow(cv.binomial.deviance)),
                        'lambda.min' = F, 'lambda.1se' = F, `lambda-1se` = F)
  #df$lambda.1se = FALSE
  #df$`lambda-1se` = FALSE

@@

```{r xvalpt, message=F, warning=F, echo=T, fig.height=16, fig.width=16}
performance.plots = list(); coefs = list(); deviance.plots = list()
plots = list()
folds = 10; splits=5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)

if (file.exists('/tmp/all.pt.alpha.Rdata')) {
  load('/tmp/all.pt.alpha.Rdata', verbose=F)
} else {
  for (a in alpha.values) {
    fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
    autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none')

@@

    plots[[as.character(a)]] = arrangeGrob(cv.patient$plot+ggtitle('CV Performance'), cv.patient$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)
  }
  save(plots, coefs, file='/tmp/all.pt.alpha.Rdata')
}

@@

cfs = as.data.frame(matrix(data=0,nrow=length(unique(names(table(unlist(lapply(coef.stable, names)))))), ncol=length(names(coef.stable)), 
                           dimnames=list( unique(names(table(unlist(lapply(coef.stable, names))))), names(coef.stable) )))

table(rowSums(cfs))

coef.stable = lapply(coef.stable, function(cf) cf/(folds*splits))

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('75%'=length(which(cf >= 0.75)),  '50%'=length(which(cf >= 0.5)),  '<50%'=length(which(cf < 0.5)))
}))
rownames(stability) = names(coef.stable)
```

Values that are closer to full lasso (alpha=1) are pretty similar, however at full lasso we see fewer features than if we regularize it at 0.8. They share all non-zero features as well.
`r pander(sapply(coef.stable, length), caption='Total features at each value of alpha')`

`r pander(table(rowSums(cfs)), caption="Number of features found in 1, 2, or all 3 models by alpha 0.8, 0.9 and 1")`

`r length(unique(unlist(lapply(coef.stable, function(cf)  names(which(cf > 0.8)) ))))` features are shared at all values of alpha and were non-zero in at least 80% of the folds.

```{r noHGD, echo=T, message=F, warning=F, fig.height=16, fig.width=16}
`%nin%` <- Negate(`%in%`)

if (file.exists('/tmp/nohgd.Rdata')) {
  load('/tmp/nohgd.Rdata', verbose=T)
} else {
  # No HGD/IMC
  samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC'))$Samplename)

  no.hgd.plots = list()
  coefs = list()
  for (a in alpha.values) {
    fit1 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
    autoplot(fit1, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
    lambda1 <- fit1$lambda

    cv.nohgd = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=lambda1, pts=subset(pts, Samplename %in% samples), a=a, nfolds=folds, splits=splits, fit=fit1, select='deviance')

    no.hgd.plots[[as.character(a)]] = arrangeGrob(cv.patient$plot+ggtitle('CV Performance'), cv.patient$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

    coef.1se = as.data.frame(non.zero.coef(fit1, cv.nohgd$lambda.1se))
    coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nohgd$non.zero.cf)
  }
  save(no.hgd.plots, coefs, file='/tmp/nohgd.Rdata')
}

do.call(grid.arrange, c(no.hgd.plots, top="No HGD/IMC samples", ncol=2))

coef.stable = lapply(coef.stable, function(cf) cf/(folds*splits))

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('75%'=length(which(cf >= 0.75)),  '50%'=length(which(cf >= 0.5)),  '<50%'=length(which(cf < 0.5)))
}))
rownames(stability) = names(coef.stable)

`r pander(stability, caption='Total features found at each value of alpha in 50% or more of the folds run')`

`r pander(table(rowSums(cfs)), justify='left')`

```{r noLGD, echo=T, message=F, warning=F, fig.height=16, fig.width=16}

if (file.exists('/tmp/nolgd.Rdata')) {
  load('/tmp/nolgd.Rdata', verbose=T)
} else {
  # No LGD
  samples = intersect(rownames(dysplasia.df), c(subset(info, Status == 'NP')$Samplename, subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD') & Status == 'P')$Samplename))

  nolgd.plots = list(); coefs = list()
  for (a in alpha.values) {
    fit2 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
    autoplot(fit2, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
    lambda2 <- fit2$lambda

    cv.nolgd = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=lambda2, pts=subset(pts, Samplename %in% samples), a=a, nfolds=folds, splits=splits, fit=fit2, select='deviance')

    nolgd.plots[[as.character(a)]] = arrangeGrob(cv.nolgd$plot+ggtitle('CV Performance'), cv.nolgd$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

    coef.1se = as.data.frame(non.zero.coef(fit2, cv.nolgd$lambda.1se))
    coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nolgd$non.zero.cf)
  }
  save(nolgd.plots, coefs, file='/tmp/nolgd.Rdata')
}

@@

coef.stable = lapply(coef.stable, function(cf) cf/(folds*splits))

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('75%'=length(which(cf >= 0.75)),  '50%'=length(which(cf >= 0.5)),  '<50%'=length(which(cf < 0.5)))
}))
rownames(stability) = names(coef.stable)

@@

`r pander(stability, caption='Total features found at each value of alpha in 50% or more of the folds run')`

@@

`r pander(table(rowSums(cfs)), justify='left')`

@@

```{r, echo=T, message=F, warning=F, fig.height=16, fig.width=16}
m = median(subset(sum.patient.data, Status == 'P')$total.samples)

@@

few.plots = list()
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, nlambda=500, family='binomial') # all patients
  cv.back.few = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=a, fit=fit0, minR=0.1, select='deviance')
  #few.plots[[as.character(a)]] = cv.back.few$plot + ggtitle(paste('alpha=',a,sep=''))

@@

  cv.back.many = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=a, fit=fit0, select='deviance')
  #many.plots[[as.character(a)]] = cv.back.many$plot + ggtitle(paste('alpha=',a,sep=''))

@@

    # Fit generated on all samples, including HGD
    a = 1
    fitLOO <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
    cv = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fitLOO$lambda, 
                                  pts=subset(pts, Samplename %in% samples), a=a, nfolds=10, splits=5, fit=fitLOO, select='deviance')
    #plots[[pt]] = cv$plot + ggtitle(pt)

    plots[[pt]] = arrangeGrob(cv$plot+ggtitle('CV Performance'), cv$deviance.plot+ggtitle('Binomial Deviance'), top=pt, ncol=2)

    fits[[pt]] = cv

@@

      coef.1se = as.data.frame(non.zero.coef(fitLOO, cv$lambda.1se))
      coefs[[pt]] = coef.stability(coef.1se, cv$non.zero.cf)

      pg.samp[[pt]]$Prediction = predict(fitLOO, dysplasia.df[pg.samp[[pt]]$Samplename,], s=cv$lambda.1se, type='response')[,1]
    } else {
      warning(paste("Patient", pt, "did not have a 1se"))
    }
  }
  save(plots, performance.at.1se, coefs, fits, pg.samp, file='/tmp/loo.Rdata')
}

# do.call(grid.arrange, c(lapply(fits[sample(subset(sum.patient.data, Status == 'NP')$Patient, 4)], function(f) {
#   plt = plot.patient.cv(f$lambdas)   
#   arrangeGrob(plt$performance +ggtitle('CV Performance'), plt$deviance+ggtitle('Binomial Deviance'), ncol=2)
# })))

# do.call(grid.arrange, c(lapply(fits[sample(subset(sum.patient.data, Status == 'P')$Patient, 4)], function(f) {
#   plt = plot.patient.cv(f$lambdas)   
#   arrangeGrob(plt$performance +ggtitle('CV Performance'), plt$deviance+ggtitle('Binomial Deviance'), ncol=2)
# })))

@@

When predicting each sample individually `r length(pg[pg == 1])` patients were correctly predicted in all samples. The remaining patients were predicted in `r max(pg[pg != 1])*100`% or fewer samples.

@@

```{r timepoints2, echo=T, message=F, warning=F}
# For all pre-HGD/IMC/LGD in progressors
pre.hgd = sapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], function(pt) {
  e = subset(pt, Pathology %nin% c('HGD', 'IMC', 'LGD'))
  e$Prediction
  sum(as.integer(e$Prediction > 0.5))/nrow(e)
})

@@

What about the non-progressors that had progressor predictions?
```{r}

np.predictions = as.data.frame(matrix(data=0,nrow=6, ncol=3, dimnames=list(c(), c('%.Predicted','n.Endoscopies','n.Patients'))))
for (i in 1:nrow(np.predictions)) {
  p = sapply(pg.samp[subset(sum.patient.data, Status == 'NP')$Patient], pred.by.endo, i)
  pp = sapply(p, predicted.progression)
  pp = pp[!is.na(pp)]
  np.predictions[i,] =  rbind(round(sum(pp[!is.na(pp)]/length(pp[!is.na(pp)])), 2), i, length(pp))
}
